---
output:
  pdf_document: default
  html_document: default
---
# Results {#Results}

This chapter presents examples demonstrating the application of semi-mechanistic models in infectious disease modeling. These examples illustrate the capabilities of semi-mechanistic models. We begin with a simulated data example to showcase the effectiveness of fitting semi-mechanistic models with a single unknown function. The structure of the compartmental models is intentionally kept simple to demonstrate that estimating unknown functions using penalized smoothers is straightforward when integrated into the syntax and optimization engine of `macpan2`. The objective is to illustrate this approach so that modelers can use it without extensive knowledge of spline and nonlinear optimization literature. Following this, we provide three real-world examples: an epidemic of scarlet fever in Ontario from 1929 to 1931, the initial SARS-CoV-19 outbreak in Ireland at the start of the pandemic, and four decades of measles in London, UK.

## SIRS Model with Simulated Data {#simulation}
One challenge in epidemiological modeling is the lack of direct observation of the transmission rate within datasets containing incidence or prevalence data. Consequently, the true shape of the function used to estimate the transmission rate remains unknown. Our goal is to develop a compartmental model, formulated as a deterministic system of ordinary differential equations, incorporating a linear smoother component. If this model can accurately predict population dynamics based on the time-varying transmission rate, it should enable us to infer the transmission rate by fitting the model to the data. This approach allows us to validate the effectiveness of the smoothing parameter estimation methodology and demonstrate the applicability of these methods with models formulated as discrete systems of ordinary differential equations.

Consider the time-varying transmission rate \(\boldsymbol{\beta}\) as a latent process. 'Time-varying' means that at each observation \(x_i\) in the dataset, the model—comprising a system of ordinary differential equations—contains an estimated value for the transmission rate for each \(1 \leq i \leq n\), where \(n\) is the number of observations.

\(\boldsymbol{\beta}\) is constructed as the linear smoother \(b_0 + \mathbf{X}\mathbf{b}\), where \(b_0\) is the intercept, \(\mathbf{X}\) is the model matrix associated with the particular smoother, and \(\mathbf{b}\) is the vector of smoothing coefficients. Both \(b_0\) and \(\mathbf{b}\) are parameters that need to be estimated. This linear smoother, which estimates \(\boldsymbol{\beta}\), is a non-parametric component inside a deterministic system of differential equations, whose trajectory is determined entirely by the starting parameters. This is why the model is called semi-mechanistic or partially specified. Some components of the model contain unknown functions, while the rest of the model comprises conventional elements with only unknown scalar parameters.

To prove that semi-mechanistic models within the `macpan2` modeling framework are capable for this problem, we conduct a simulation study. For a chosen smooth, we simulate data and add Gaussian noise to the incidence. The simulated data is used to calibrate the model using various smoothers available in the `mgcv` package. In Figure \ref{fig:sim_basis}, we illustrate the shape of the basis functions for each of the univariate smoothing bases used in this methodology.

Consider a SIRS compartmental model with a fixed waning immunity parameter \(\phi\). The total number of individuals in the population is given by \(N\). The initial values of \(S\), \(I\), and \(R\) are fixed at the endemic equilibrium solutions.

We construct the data-generating model as follows. The smoother type and the number of knots \(k\) are specified using a particular `mgcv` smooth. This determines the form of \(\mathbf{X}\) and the penalty matrix \(\mathbf{P}\). The smoothing coefficients \(\mathbf{b}\), of dimension \(k-2\), are assumed to be multivariate Gaussian. Thus, \(\mathbf{b}\) is defined as random normal deviates at the \(k-2\) evenly spaced quantiles with a mean of 0 and a standard deviation specified as \(b_{sd}\). As the variation of this distribution increases, the true function describing the time-varying transmission rate for the data model will become more complex. An initial value for \(b_0\) is chosen as the log of the initial value of \(\beta\). The recovery rate \(\gamma\) is fixed. Initial values for the remaining parameter estimates are set to . 

The model trajectory is simulated from these initial conditions using Euler steps. Gaussian noise (\(sd = 0.2\)) is added to the log-transformed simulated incidence vector, and then the inverse transformation is applied. This vector is used to test the efficacy of the semi-mechanistic models. The models are calibrated using `macpan2`, which uses the Laplace approximation via `TMB` to optimize the objective function and update parameter estimates using Quasi-Newton methods.

To investigate the effect of temporal aggregation on model performance, we conducted an additional study where we aggregated the simulated data on a weekly scale before fitting the model. This process involved summing the predicted incidence over each week and then fitting the model to these aggregated observations. Both the unaggregated and aggregated models use the same simulated trajectory with added noise, ensuring consistency between the two approaches. We used \(k=20\) knots to simulate the data. The aggregated data was obtained by summing the noisy unaggregated data on a weekly basis.

We performed cross-model comparison by varying the smoothing basis used to calibrate the models across the smoothing basis used to simulate the data. The variance used to generate the starting values for the smoothing coefficients \(b_{sd} = 1\) is fixed for starting values. \(\gamma = 1/7\), which represents a one-week period of an infectious individual being capable of transmitting the disease to a susceptible individual. \(\phi = \frac{1}{300}\), which means that at each time point, that proportion of the recovered individuals lose immunity. An equivalent interpretation of the waning immunity parameter is that it represents the period of immunity for an individual, which in this case represents 300 days. We have kept \(\gamma\) fixed in the simulation study, but in the real-life data examples, we have used a log-normal prior to allow flexible deviation from the initialized starting value of \(\gamma\).

In Figures \ref{fig:sim_gp}, \ref{fig:sim_tp}, \ref{fig:sim_cr}, and \ref{fig:sim_bs}, we present the results for unaggregated data simulated using the Gaussian process (GP), thin plate regression spline (TP), cubic regression spline (CR), and B-spline (BS) bases, respectively. In Figures \ref{fig:sim_agg_gp}, \ref{fig:sim_agg_tp}, \ref{fig:sim_agg_cr}, and \ref{fig:sim_agg_bs}, we present the results for aggregated data. We have chosen to omit the results of the calibration models utilizing a P-spline basis, as they show poor ability to predict incidence and infer the underlying true transmission rate when the number of knots used is less than the number used to generate the data.

For the predicted incidence, the red line represents the true value from the simulated trajectory with Gaussian noise. The blue line indicates the predicted incidence, with light and dark blue bands representing the 95% and 50% confidence intervals, respectively.

For the estimated transmission rate (\(\beta_t\)), the red line represents the true transmission rate function from the simulated data. The green line indicates the estimated transmission rate, with light and dark green bands representing the 95% and 50% confidence intervals, respectively.

For the estimated effective reproduction number \(R_t\), the black line represents the true reproduction number from the simulated data. The purple line indicates the estimated reproduction number, with light and dark purple bands representing the 95% and 50% confidence intervals, respectively.

The number of knots used to calibrate the models was \(k =20\) for simulated data sets, except the TP basis for both the unaggregated and aggregated data. For the GP, CR, and BS data, the value \(k =20\) was chosen as the smallest value of \(k\) to fit the predicted and estimated values without introducing overfitting. This was evaluated by visually inspecting the results of the figures below. For the TP basis, the optimization procedure for the model calibration did not converge for values of \(k \geq 10\). None of the bases were able to match the highly oscillating trajectory observed in the last 75 days of Figures \ref{fig:sim_tp} and \ref{fig:sim_agg_tp}.

In Tables \ref{tab:aic-table-sim} and \ref{tab:aic-table-sim-agg}, we present a comparison of the conditional AIC scores for all sixteen models for the unaggregated and aggregated data sets, respectively. We observe that the cubic regression spline (CR) model consistently has the lowest AIC score across all datasets. A lower conditional AIC score indicates that the model is relatively the best fit for the data, taking into account the effect of penalization on the number of knots used, i.e., effective degrees of freedom. The best model is the most likely to minimize information loss when approximating the true data-generating process, compared to the other models tested. The AIC score, which balances model fit and complexity, suggests that the CR model provides the best trade-off between these two aspects among the models compared. However, the \(\Delta\text{AIC}\) values are within two points of each other, indicating that this difference is negligible and one model is not significantly better than the others.

In Tables \ref{tab:mse-table-sim} and \ref{tab:mse-table-sim-agg}, we present a comparison of the mean square error (MSE) for all sixteen models for the unaggregated and aggregated data sets, respectively. For the unaggregated data, we observe that models using the same basis for generation and fitting have the lowest MSE for all bases except the B-spline basis. For the aggregated data, the models using the same basis for generation and fitting have the lowest MSE for all bases except the thin plate spline basis.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/unaggregated/simulation_gp_20_k(5,10,20)_bsd1_beta1_plot_basis.png}
\caption[Basis Functions for Smoothing Basis.]{\textbf{Basis functions for calibrating smoothers}. Basis matrices are obtained via mgcv::smoothCon() with number of knots \(k=20\) and domain of \(n=200\) days. Each basis matrix is determined only by the input domain and the number of knots. Each basis function is a single column of the basis matrix. The acronyms for the smoothers are defined as follows: "gp" = Gaussian process, "tp" = thin plate regression spline, "cr" = cubic regression spline and "bs" = B-spline.}
\label{fig:sim_basis}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/unaggregated/sim_combined_gp.png}
\caption[SIRS model with simulated data (GP).]{\textbf{SIRS Model with Simulated Data (GP).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. \(k=20\) knots were used to calibrate the model.}
\label{fig:sim_gp}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/unaggregated/sim_combined_tp_k8.png}
\caption[SIRS model with simulated data (TP).]{\textbf{SIRS Model with Simulated Data (TP).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. \(k=9\) knots were used to calibrate the model.}
\label{fig:sim_tp}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/unaggregated/sim_combined_cr.png}
\caption[SIRS model with simulated data (CR).]{\textbf{SIRS Model with Simulated Data (CR).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. \(k=20\) knots were used to calibrate the model.}
\label{fig:sim_cr}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/unaggregated/sim_combined_bs.png}
\caption[SIRS model with simulated data (BS).]{\textbf{SIRS Model with Simulated Data (BS).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. \(k=20\) knots were used to calibrate the model.}
\label{fig:sim_bs}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/aggregated/sim_agg_combined_gp.png}
\caption[SIRS model with aggregated data (GP).]{\textbf{SIRS Model with Aggregated Data (GP).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. The data was simulated on a daily scale and then aggregated to a weekly scale for model calibration. \(k=20\) knots were used to calibrate the model.}
\label{fig:sim_agg_gp}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/aggregated/sim_agg_combined_tp_k10.png}
\caption[SIRS model with aggregated data (TP).]{\textbf{SIRS Model with Aggregated Data (TP).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. The data was simulated on a daily scale and then aggregated to a weekly scale for model calibration. \(k=9\) knots were used to calibrate the model.}
\label{fig:sim_agg_tp}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/aggregated/sim_agg_combined_cr.png}
\caption[SIRS model with aggregated data (CR).]{\textbf{SIRS Model with Aggregated Data (CR).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. The data was simulated on a daily scale and then aggregated to a weekly scale for model calibration. \(k=20\) knots were used to calibrate the model.}
\label{fig:sim_agg_cr}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{figure/Simulated/aggregated/sim_agg_combined_bs.png}
\caption[SIRS model with aggregated data (BS).]{\textbf{SIRS Model with Aggregated Data (BS).} The columns represent the predicted incidence, the estimated transmission rate and effective reproduction number. The rows correspond to the different smoothing basis used to fit the model to the data. The data was simulated on a daily scale and then aggregated to a weekly scale for model calibration. \(k=20\) knots were used to calibrate the model.}
\label{fig:sim_agg_bs}
\end{figure}



```{r aic-table-sim, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(tidyr)

# Load the saved R object
load("data/Tables/Simulated/unaggregated_table.RData")

# Generate and display the table with additional headers
# Generate and display the table with the revised caption
kable(final_result, format = "latex", booktabs = TRUE, 
      caption = "\\textbf{Conditional AIC Scores for SIRS Model with Simulated Data}. The columns represent the smoothing basis used to fit the model, while the rows indicate the basis used to generate the simulated data. \\(\\Delta \\text{cAIC}\\) values are calculated relative to the best score within each row.") %>%
  kable_styling(latex_options = c("hold_position"))

```

```{r mse-table-sim, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(tidyr)

# Load the saved R object
load("data/Tables/Simulated/MSE_unaggregated.RData")


# Generate and display the table with the revised caption
kable(adjusted_df, format = "latex", booktabs = TRUE, 
      caption = "\\textbf{Mean Square Error (MSE) for SIRS Model with Simulated Data}. The columns represent the smoothing basis used to fit the model, while the rows indicate the basis used to generate the simulated data.  \\(\\Delta \\text{MSE}\\) values are calculated relative to the best score within each row.") %>%
  kable_styling(latex_options = c("hold_position"))

```

```{r aic-table-sim-agg, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(tidyr)

# Load the saved R object
# Load the data
load("data/Tables/Simulated/aggregated_table.RData")



# Generate and display the table with the revised caption
kable(final_result, format = "latex", booktabs = TRUE, 
      caption = "\\textbf{Conditional AIC Scores for SIR Model with Simulated Data Aggregated to a Weekly Scale}. The columns represent the smoothing basis used to fit the model, while the rows indicate the basis used to generate the simulated data. The trajectories were simulated on a daily scale and then aggregated to a weekly scale for model calibration. \\(\\Delta \\text{cAIC}\\) values are calculated relative to the best score within each row.") %>%
  kable_styling(latex_options = c("hold_position"))
```

```{r mse-table-sim-agg, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(tidyr)

# Load the saved R object
# Load the data
load("data/Tables/Simulated/MSE_aggregated.RData")
# Remove row names
rownames(final_result) <- NULL

# Generate and display the table with the revised caption
kable(final_result, format = "latex", booktabs = TRUE, 
      caption = "\\textbf{Mean Square Error (MSE) for SIR Model with Simulated Data Aggregated to a Weekly Scale}. The columns represent the smoothing basis used to fit the model, while the rows indicate the basis used to generate the simulated data. The trajectories were simulated on a daily scale and then aggregated to a weekly scale for model calibration. \\(\\Delta \\text{MSE}\\) values are calculated relative to the best score within each row.") %>%
  kable_styling(latex_options = c("hold_position"))
```


## Scarlet Fever in Ontario 1929-1931 {#scarlet}
We consider an epidemic of scarlet fever in Ontario from 1929 to 1930. The dataset comes from the International Infectious Disease Data Archives (IIDDA) [@walkerCanmodIiddaInternational2024].

The calibration model is constructed using the basic SIR model, with an initial recovery rate of \(\gamma = 1/7\) and an initial number of infected of \(I_0 = 50\).

In Figure \ref{fig:scarlet_combined}, we show the predicted incidence, estimated transmission rate, and effective reproduction number for the best model, which uses \(k = 9\). This model employs the Ornstein-Uhlenbeck (OU), thin plate regression spline (TPRS), cubic regression spline (CR), and cyclic cubic regression spline (CC) bases for the linear smoother component. The power OU kernel for the GP basis was chosen to be consistent across all the real data examples. We observed little difference when varying the kernel type.

The shape of the estimated transmission rate and reproduction number varies across the smoothing bases for this example. The GP basis shows a positive slope as the rate of transmission reaches its maximum, whereas the TPRS, CR, and CC bases start at or very near the maximum.

In Table \ref{tab:aic-table-scarlet}, we present the conditional AIC scores and MSE calculations for the four models. We provide the delta values, which are calculated by subtracting the best score from each result for both the AIC and MSE calculations. The AIC score for the cubic regression spline (CR) basis is the lowest compared to the other bases. However, the difference in AIC scores between the bases is less than 2, implying that no single model should be chosen over the others based solely on the AIC. In the same table, we observe that the mean square error (MSE) is lowest for the thin plate spline (TP) model rather than the CR model.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth, height = \textwidth]{figure/Scarlet/scarlet_combined.png}
\caption[Combined Analysis of Scarlet Fever (1929-1930)]{\textbf{Combined analysis of predicted incidence, estimated transmission rate, and effective reproduction number for weekly observed scarlet fever cases in Ontario, from 1929 to 1930.} Each column represents the predicted incidence, estimated transmission rate, and effective reproduction number, respectively. Each row corresponds to different smoothing bases used to calibrate the data, all fitted with \(k=20\) knots. The true incidence is shown in red. The covariance function of the GP basis uses a power exponential kernel, with power parameter \(\kappa = 1\) and range parameter \(\ell = 2\). The figures display 95\% and 50\% confidence intervals.}
\label{fig:scarlet_combined}
\end{figure}

  

```{r aic-table-scarlet, echo=FALSE, message=FALSE, warning=FALSE,}
library(knitr)
library(kableExtra)
library(tidyr)

# Load the saved R object
load("data/Tables/ScarletFever/MSE_AIC_Scarlet.RData")

# Generate and display the table with additional headers
# Generate and display the table with additional headers
kable(combined_df, format = "latex", booktabs = TRUE, 
      col.names = c("Basis Type", "Delta MSE", "Delta cAIC"),
      caption = "\\textbf{\\(\\Delta \\text{cAIC}\\) and \\(\\Delta \\text{MSE}\\) Scores for Ontario Scarlet Fever (1929-1930).} \\(\\Delta \\text{cAIC}\\) and \\(\\Delta \\text{MSE}\\) values are calculated relative to the best score.") %>%
  kable_styling(latex_options = c("hold_position", "float_placement"))

```

## Covid-19 in Ireland 2020 {#Ireland}
Next, we fit the model to observations of the daily number of COVID-19 cases in Ireland from the onset of the outbreak, spanning February 20, 2020, to May 9, 2020. This data is sourced from the publication by Andrade and Duggan [@andradeInferringEffectiveReproductive2022].

The calibration model is constructed using the basic SIR model, with an initial recovery rate of \(\gamma = 1/6\), which is obtained from the results of Park et al. [@parkImportanceGenerationInterval2022].

Figure \ref{fig:ireland_combined} presents the results of the models using the GP, TPRS, and CR smoothing bases. The BS basis functions exhibited extremely large uncertainty estimates, so we did not include the results. All bases displayed an excellent fit to the observed incidence data with appropriately sized uncertainty estimates.

Table \ref{tab:aic-table-ireland} provides the delta values for the conditional AIC scores and the MSE calculations. The CR basis has the best AIC score, while the GP basis has the lowest MSE calculation. The difference in AIC scores between the model with the CR basis and those with the GP and TP bases is less than two, suggesting that the AIC score alone cannot distinguish the best basis for fitting the data.

The model's sensitivity to changes in the starting value of \(\gamma\) (ranging from 7 to 14 days) does not alter the shape of the functional form of \(\beta\), but it changes the amplitude of the peak, leading to an increase in \(R_t\). Increasing the variance of the log-normal prior on \(\gamma\) causes the model to tend to select a larger \(\gamma\). However, this also results in a dramatic increase in uncertainty estimates. We observed that when the starting value of \(\gamma\) is lowered and the variance of the log-normal prior is increased simultaneously, the model still predicts larger values of \(\gamma\).

Andrade and Duggan (2022) used COVID-19 data to infer the effective reproduction number by employing an SEIR model with three different data-generating processes for the transmission rate. One of the processes they used was Geometric Brownian Motion (GBM). GBM is similar to the Ornstein-Uhlenbeck (OU) process that we used, but it is non-stationary, non-mean-reverting, and the response is log-normally distributed. In our thesis, we implemented the OU process as a special case of a Gaussian Process (GP) with an exponential covariance function. By viewing the OU process as a GP, we focus on the joint distribution of values at different times, characterized by the mean and covariance functions. These functions describe the decay rate of the covariance and the process variance, respectively, emphasizing the correlation structure. Computationally, this approach allows us to use linear algebra. Andrade and Duggan define the transmission rate using GBM formulated as an SDE. They also assume that the response, the incidence data, follows Poisson and Negative Binomial distributions. Additionally, they use Apple mobility data to adjust the transmission rate by assuming that the effect of social distancing is correlated with the transmission rate.

We compared our results, using the OU process basis for our linear smoother, with those of Andrade and Duggan. They aggregated their incidence data to a weekly scale to account for irregularities in daily reporting. We implemented this by computing the trajectory on a daily scale, aggregating the predicted incidence to a weekly scale, and then fitting the model to the weekly aggregated observed data. The daily trajectory of the transmission rate and reproduction number estimates are then averaged over each week.

The predicted incidence fits the data well, with very reasonable uncertainty bounds, and the shape is very similar to Andrade and Duggan's results. The estimated transmission rates, ignoring the first week in our plots, both show approximately the same shape, an exponential decay. The magnitude of the transmission rate is about twice as large for Andrade and Duggan compared to our results. The estimated effective reproduction number, which is the transmission rate scaled by the recovery rate and the proportion of susceptibles at any time \(t\), also shows a similar pattern. Although Andrade and Duggan compute an analytical expression for the basic reproduction number and we compute ours as the transmission rate scaled by the recovery rate, we both scale the basic reproduction number by the ratio \(\frac{S_t}{N_t}\) to obtain the effective reproduction number. Most notably, we observe that the estimated effective reproduction number has not only the same shape as Andrade and Duggan's but also a similar magnitude. The uncertainty estimates are tighter in our model, but our model is simpler.


```{r aic-table-ireland, echo=FALSE, message=FALSE, warning=FALSE,}
library(knitr)
library(kableExtra)
library(tidyr)

# Load the saved R object
load("data/Tables/Ireland/MSE_AIC_Ireland.RData")

# Generate and display the table with additional headers
kable(combined_df, format = "latex", booktabs = TRUE, 
      col.names = c("Basis Type", "Delta MSE", "Delta cAIC"),
      caption = "\\textbf{\\(\\Delta \\text{cAIC}\\) and \\(\\Delta \\text{MSE}\\) Scores for Calibrating Models with Varying Smoothing Basis, Calibrated to Ireland Covid-19 (2020).} \\(\\Delta \\text{cAIC}\\) and \\(\\Delta \\text{MSE}\\) values are calculated relative to the best score.") %>%
  kable_styling(latex_options = c("hold_position", "float_placement"))

```

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth, height = \textwidth]{figure/Ireland/ireland_combined.png}
\caption[Combined Analysis of Covid-19 in Ireland (2020)]{\textbf{Combined analysis of predicted incidence, estimated transmission rate, and effective reproduction number for observed Covid-19 cases in Ireland, 2020.} The data is given on a daily scale, and the trajectory is simulated daily but aggregated to a weekly scale for fitting the model. Each column represents the predicted incidence, estimated transmission rate, and effective reproduction number, respectively. Each row corresponds to different smoothing bases used to calibrate the data, all fitted with \(k=7\) knots. The true incidence is shown in red. The covariance function of the GP basis uses an exponential kernel function, which makes the basis an Ornstein-Uhlenbeck process, with exponent parameter \(\kappa = 1\) and range parameter \(\ell = 2\). The figures display 95\% and 50\% confidence intervals.}
\label{fig:ireland_combined}
\end{figure}

## Measles London UK 1944-1984 {#Measles}
We now present a more challenging problem than the previous examples. We consider weekly observed measles cases in London, UK, from 1944 to 1984. This dataset was first utilized in a publication by David Earn et al. [@earnSimpleModelComplex2000]. 

The calibration model is constructed using the basic SIR model. The starting value for the recovery rate is \(\gamma = 1/8\), aligning with the initial value used in [@earnSimpleModelComplex2000]. A key assumption we make is that the total population remains constant over time. This is due to the nature of the SIR model, which does not account for a time-varying total population component. We fixed the total population to \(N = 8,615,050	\), which is the population of London at the start of this dataset. This assumption is reasonable, as the population of London has fluctuated between six and ten million from then until the present day. However, since the disease predominantly affects children, a more sophisticated model could include a time-varying total population component, partitioned according to age and weighted by the incidence rate per age demographic. We initialize the number of infected individuals at \(I_0 = 250\). 

The full dataset extends to 1994, but we had difficulty tuning the model to fit the last decade. During this period, the observed incidence was relatively flat compared to the previous years. We observed that the calibrated model inflated the predicted transmission rate to unreasonable levels. By truncating the last decade, reducing the observations from 2,660 to 2,140, we were able to calibrate the model more effectively.

We exclusively present the results for a Gaussian process basis for the linear smoother component of the model. Other smoothing bases were tested, but the optimizer failed to converge for more than 100 to 200 observations, as the Newton-Raphson method failed to find a minimum. In Figure \ref{fig:Measles_trans}, we show the predicted incidence, transmission rate, and effective reproduction number for the SIR model with a Gaussian process basis for the linear smoother component. The covariance function is the exponential kernel with a range parameter \(\ell = 20\). Other kernels, such as the Matérn function, were tested by iterating over different range parameters \(\ell = 30, 40, 50\). The resulting models showed little difference in optimized parameter values and conditional AIC scores. For simplicity, we present the parsimonious model with the simplest kernel function and the smallest range parameter. Each model took about 20 minutes to fit.

Figure \ref{fig:Measles_trans} displays the optimized values of the parameters in the calibrated model and their uncertainty measurements. These are the log transformed values. Exponentiating, the partial prior on \(\gamma\) produced an optimal value of \(\frac{1}{11.84}\). 

\begin{figure}
\centering
\includegraphics[width=\textwidth, height = \textwidth]{figure/Measles/Measles_combined_plot.png}
\caption[Combined Analysis of Measles in London, UK, (1944-1984)]{\textbf{Predicted incidence and estimated transmission rate and effective reproduction number for weekly observed measles cases in London, UK, from 1944 to 1984 using a Gaussian Process smoother.} The linear smoother component of the semi-mechanistic compartmental model employs an Ornstein-Uhlenbeck (OU) process with a range parameter \(\ell = 20\) and \(k = 100\) knots. The model estimates a recovery rate of approximately \(\gamma = 11.84\) days and an initial number of infected individuals of approximately \(I_0 = 250\). The figures display 95\% and 50\% confidence intervals.}
\label{fig:Measles_trans}
\end{figure}






